Quality Assesment of OpenStreetMap Footpath Data

### 2020-06-17 report/report.tex
Use overlap instead of collision ouside of techincal contexts

To examine this, a slice of the OSM dataset in the Stockholm area has been taken, and each road has been assigned with standard withds according to swedish regulation, after which a geometrical algorithm was applied to all road pairs to obtain the average amount of road overlap.
It was found that footpaths were less problematic than other road types (primary, secondary, residential), in terms of feature collisions, and thus should pose no more problems in 3D city reproduction than other road types.

@@

% TODO: Use overlap instead of collision outside of technical contexts

@@

In this project it was hypothised that the risk of feature collision may be higher between streets and roads designated for energy efficient mobility, such as footpaths, since they are much more organic and dense in their nature than main streets.
Thus, it is often in the designer's best interest to have a procedure for importing road data that results in minimal feature overlap, while still retaining accuracy to the real world.
A possible way of mitigating this problem is looking towards gathering geodata from volunteer-based geoservices such as OSM.
The OSM project is a freely available wiki-like geodatabase based on Volunteered Geographical Information (VGI), meaning anyone can contribute by adding or editing features.
The OSM world dataset is being consistently updated by volonteers from all over the world, and the main idea is that by using OSM geodata for city reproduction, the temporal aspect of data accuracy would in theory be more assured by the crowd contribution nature.
There are already a few such services that use OSM data to generate 3D geodata visualizations for media and video game applications.
A few prominent examples are Mantle
\footnote{Mantle homepage. Accessed 17-05-2020. https://www.mantle.tech/}
and Mapbox
\footnote{Mapbox homepage. Accessed 17-05-2020. https://www.mapbox.com/}.
Such services are however usually limited in the accuracy and type of data that they can provide.
While they can provide both building footprints and road networks, they usually provide road data with no respect to overlap between features.

@@

For the primary research question: the collision study, a program has taken a slice of the OSM dataset in the Stockholm area and assigned each road with standard withds according to swedish regulation.
It has then identified critical areas in the OSM dataset, where the paths are so wide that they collide with existing map features.
It has also identified how many of those areas can in theory be corrected by simple translation of the road vertices, without propagation of collision to other features.
Here, features are defined as the geometrical components that comprise the OSM map.
For the purpose of this report, features will refer mainly to roads and building footprints.
For the collision study, see sections 3.1.2 for an implementation overview, 3.3 for a detailed walkthrough of the program implementation, and 4.1.1 for the results.

For the secondary research question: the geodata precision study, the positional accuracy will be estimated by comparing feature points between the OSM building dataset and the Swedish University of Agricultural Sciences (SLU) building datset with known accuracy. See section 3.4.2 for an overview of the SLU dataset and how it is obtained.
The positional accuracy will be needed to assess where road collisions occur and whether they can easily be corrected.
For the positional accuracy study, see sections 3.1.1 for an implementation overview, 3.2 for a walkthrough of the program implementation, and 4.1.2 for the results.

@@

This chapter will dive into the state-of-the-art use of city reconstruction methods within the research fields of visualization and urban design.
This chapter will present the motivations for the digitization of urban planning, as well as a brief history of the field of OpenStreetMap data qualification, and furthermore present the research papers whose methods are at the foundation of this project.
Lastly, it will present the evaluation metrics that this project served to obtain, and how they relate to the eventual collision analysis.
For the geodata precision study specifically, it will examine how geodata is qualifyied in general, by examining van Oorts quality criteria, and how these will be used in this project.

@@

The primary study of the rate of collision between roads and other features will serve to obtain the following three metrics:

@@

The road features will be categorized according to type, so that the rate of collisions between footpaths and other features can be compared to that of other road types.
OSM mainly makes the distinction between primary, secondary, residential and footpath roads, so this classification will also be used in this report.
The above features will be calculated for each road type separately by summing over each road feature of a certain type.
Following this, an algorithm will find the exact same metrics with regards to how many road features that can in theory be corrected by simple geometrical translation of their vertices.
This will also be preseneted in terms of the number of road features, the number of edges and the total length in meters of remaining colliding road.

@@

This chapter will provide a technical overview of the data collection process and the algorithms used for the purpose of this study, starting with a brief overview in section 3.1.
Section 3.2 will present the data collection process for the geodata precision study. It will present how to qualify dataset completeness and building correspondence (how many of the buildings in one dataset that can be found in the other) by collecting the cumulative building areas and looking at what buildings have overlapping areas between the two datasets.
The definitions and algorithms neccessary for calculating posiitonal accuracy and a measure for shape accuracy will also be presented.
Section 3.3 will present at the data collection process for the collision study. It will present how standard road withs were obtained, as well as the algorithm for detemining feature collision that was used.
Section 3.4 will discuss briefly the data sources used in this project, why they were chosen and specific issues with the data collection process.

@@

Programmatic collection of statistics has been neccessary for both the primary and secondary research questions, and this section presents a broad overview of the technical implementation behind the map analysis and data collection processes for both the geodata precision study and the collision study.

@@

\subsection{Collision study}

Once the positional error estimate was obtained from the precision study, it was used to assess the integrity of the OSM Stockholm road network.
This was done by firstly locating critical points where assigning standard widths to the roads would cause collision with other features, and then evaluating how many of those points could be eliminated by making ajustments within the found margin of error.
For the purpose of the collision study, each road has been assigned a standard width that is in line with Swedish regulations, since in the OSM dataset, roads are represented as simple polylines, lacking width.
After the width assignment, a program iterated over all features and calculated how many of them that show overlap with any other feature, as well as the total length of the road polylines that is intersecting some other feature.
Finally, the program calculated how many of these colliding features that can be corrected by making adjustments that are within the positional accuracy of the OSM data, as obtained by the geodata precision study.

@@

Finally, when two building footprints and their vertices have been mapped, the average offset of matching vertices will be used as a measure of positional error.
The average and variance in vertex spread will be calculated and used to make an assessment of the geometric precision of the OSM dataset, and this precision will further be used to form an upper boundary on geometrical corrections that can be made in the collision study.

\section{Collision study}

@@

The standard road widths were obtained from a publication by the Swedish Transport Administration
\footnote{Vägars och gators utformning - Trafikverket. Publikation 2020:029. https://trafikverket.ineko.se/Files/sv-SE/71830/\\Ineko.Product.RelatedFiles/2020\_029\_vagar\_och\_gators\_utformning\_krav.pdf}
, and were presented as requirements on the minimum lane width for certain types of roads and applications.
These regulated minimum widths were used as a lower bound road with when examining the OSM dataset for potential collisions.
The upper bound was found by visual measurements of a few key locations in the Stockholm area using the Bing maps service.
Table \ref{table:road-widths} shows the minimum and maximum road widths that will be used in the collision study.

\begin{table}[H]
    \begin{tabular}{lllll}
                      & Footpath & Residential & Secondary & Primary \\
        Minimum width & 2.0 m    & 6.0 m       & 6.5 m     & 8.0 m   \\
        Maximum width & 5.0 m    & 9.0 m       & 10.0 m    & 16.0 m
    \end{tabular}
    \captionof{table}{Standard road widths used for the collision study}
    \label{table:road-widths}
\end{table}

@@

The algorithm begins by iterating over all road features.
Let $A$ be the current road feature.
Since only road collisions are of interests, all roads were added to a queue and gone over one by one.
The polyline of the current road feature is extracted.
Then, all other candidate features for collision are iterated over, and their boundary polygons (for buildings) or polylines (for roads) are extracted.
Let $B$ be the current feature, road or building, that is being compared to $A$.
The algorithm then iterates over each edge $e_a$ in the polyline of $A$, and each edge $e_b$ in the polyline or polygon of $B$.
The Projected Line Segment Distance algorithm (See section 4.2.3) is used to determine the shortest distance between $e_a$ and $e_b$, and the smallest distance between any combination of edges $e_a$ and $e_b$ is taken to be the shortest distance between features $A$ and $B$.
If this distance is smaller than the minimum width of $A$, then $A$ and $B$ are colliding.

To speed up the collision checking process, a geometric hashing solution was developed.
In this way, feature collision lookup could be made without comparing a feature to all other available features, but instead from a selection of the closest candidates.
The hash bucket size was 230 by 230 meters and a feature could be in multiple hash buckets.

@@

Here the results of the statistics collections will be presented.
First the results of the collision study is presented as how many of the road features of each type display collision with some other feature.
Then the results of the geodata precision is presented.
The final estimate of the positional accuracy of the OSM dataset will be presented, as well as the dataset completeness and statistics about building correspondence and shape accuracy.

@@

For the primary collision study.
Table \ref{table:road-collision} shows the statistics of the collision study, and the number of roads, number of edges and total edge length that showed intersections with other features, by road category.
For footpaths in particular, it was shown that 12.05\% of all roads collide with some other feature after extrusion to the minimum width.
The same was shown to be true for 0.11\% of the road polyline edges and 0.09\% of the total road length.
Table \ref{table:road-collision-corrected} shows the remaining roads, edges and total edges length whose intersections with other features could not be resolved by geometrical translation of a distance smaller than the positional accuracy.
For footpaths this was shown to hold for 0.51\% of all roads, 
The number of road edges and percentage of the total road length that could not be corrected by geometric translation was significantly small.

@@

    \end{tabular}
    \captionof{table}{The first result table of the collision study. This table shows the amount of features and edges, as well as the cumulative edge length, in total and which are intersecting with any other feature.}
    \label{table:road-collision}
\end{table}

@@

    \end{tabular}
    \captionof{table}{The second result table of the collision study. This table shows the amount of features and edges, as well as the cumulative edge length, which can be corrected by simple geometric translation.}
    \label{table:road-collision-corrected}
\end{table}

@@

Now that the colliison metrics have been obtained, conclusions may be drawn about the technical feasibility of 3D city reproduction from OSM data.
First the geodata precision study will be discussed.
The outcome of that study will be used to assess the quality of the positional accuracy estimate by looking at the dataset completeness, the low semantic accuracy and the unique characteristics of the Stockholm area.
Secondly the collision study will be discussed.
The main outcome is that footpaths will not be more problematic than other road types in city reproduction from OSM data.
This claim will be motivated with the data obtained in section 4.1.
Finally, the character of the feature collision cases will be discussed and the most common cases will be identified.
The chapter will conclude with a presentation of geometrical algorithms for how to correct each case automatically.
For conclusions about the usage of OSM city reproduction in engineering practise and remaining challenges, see section 5.

@@

The immediate conclusion of this study is that, in terms of feature collision, footpaths were not significantly worse than other road categories.
As a result, it can promptly be assumed that city modelling services which already utilize OSM car road data could also use OSM footpath data without a drastic increase in feature collision.
The biggest reason for this is thought to be that footpaths in their nature are narrow, and so narrow in fact that they can be displaced by more than half of their full width (minimum 1 meter) before the limit of half the positional accuracy (1.025 meters) is reached.

>>>>

To examine this, a slice of the OSM dataset in the Stockholm area has been taken, and each road has been assigned with standard withds according to swedish regulation, after which a geometrical algorithm was applied to all road pairs to obtain the average amount of road overlap.
It was found that footpaths were less problematic than other road types (primary, secondary, residential), in terms of feature overlap, and thus should pose no more problems in 3D city reproduction than other road types.

@@

@@

In this project it was hypothised that the risk of feature overlap may be higher between streets and roads designated for energy efficient mobility, such as footpaths, since they are much more organic and dense in their nature than main streets.
Thus, it is often in the designer's best interest to have a procedure for importing road data that results in minimal feature overlap, while still retaining accuracy to the real world.
A possible way of mitigating this problem is looking towards gathering geodata from volunteer-based geoservices such as OSM.
The OSM project is a freely available wiki-like geodatabase based on Volunteered Geographical Information (VGI), meaning anyone can contribute by adding or editing features.
The OSM world dataset is being consistently updated by volonteers from all over the world, and the main idea is that by using OSM geodata for city reproduction, the temporal aspect of data accuracy would in theory be more assured by the crowd contribution nature.
There are already a few such services that use OSM data to generate 3D geodata visualizations for media and video game applications.
A few prominent examples are Mantle
\footnote{Mantle homepage. Accessed 17-05-2020. https://www.mantle.tech/}
and Mapbox
\footnote{Mapbox homepage. Accessed 17-05-2020. https://www.mapbox.com/}.
Such services are however usually limited in the accuracy and type of data that they can provide.
While they can provide both building footprints and road networks, they usually provide road data with no respect to overlap between features.

@@

For the primary research question: the road overlap study, a program has taken a slice of the OSM dataset in the Stockholm area and assigned each road with standard withds according to swedish regulation.
It has then identified critical areas in the OSM dataset, where the paths are so wide that they collide with existing map features.
It has also identified how many of those areas can in theory be corrected by simple translation of the road vertices, without propagation of collision to other features.
Here, features are defined as the geometrical components that comprise the OSM map.
For the purpose of this report, features will refer mainly to roads and building footprints.
For the overlap study, see sections 3.1.2 for an implementation overview, 3.3 for a detailed walkthrough of the program implementation, and 4.1.1 for the results.

For the secondary research question: the geodata precision study, the positional accuracy will be estimated by comparing feature points between the OSM building dataset and the Swedish University of Agricultural Sciences (SLU) building datset with known accuracy. See section 3.4.2 for an overview of the SLU dataset and how it is obtained.
The positional accuracy will be needed to assess where road overlap occurs and whether the collisions can easily be corrected.
For the positional accuracy study, see sections 3.1.1 for an implementation overview, 3.2 for a walkthrough of the program implementation, and 4.1.2 for the results.

@@

This chapter will dive into the state-of-the-art use of city reconstruction methods within the research fields of visualization and urban design.
This chapter will present the motivations for the digitization of urban planning, as well as a brief history of the field of OpenStreetMap data qualification, and furthermore present the research papers whose methods are at the foundation of this project.
Lastly, it will present the evaluation metrics that this project served to obtain, and how they relate to the overlap analysis.
For the geodata precision study specifically, it will examine how geodata is qualifyied in general, by examining van Oorts quality criteria, and how these will be used in this project.

@@

The primary study of the rate of overlap between roads and other features will serve to obtain the following three metrics:

@@

The road features will be categorized according to type, so that the rate of overlap between footpaths and other features can be compared to that of other road types.
OSM mainly makes the distinction between primary, secondary, residential and footpath roads, so this classification will also be used in this report.
The above features will be calculated for each road type separately by summing over each road feature of a certain type.
Following this, an algorithm will find the exact same metrics with regards to how many road features that can in theory be corrected by simple geometrical translation of their vertices.
This will also be preseneted in terms of the number of road features, the number of edges and the total length in meters of remaining colliding road.

@@

This chapter will provide a technical overview of the data collection process and the algorithms used for the purpose of this study, starting with a brief overview in section 3.1.
Section 3.2 will present the data collection process for the geodata precision study. It will present how to qualify dataset completeness and building correspondence (how many of the buildings in one dataset that can be found in the other) by collecting the cumulative building areas and looking at what buildings have overlapping areas between the two datasets.
The definitions and algorithms neccessary for calculating posiitonal accuracy and a measure for shape accuracy will also be presented.
Section 3.3 will present at the data collection process for the overlap study. It will present how standard road withs were obtained, as well as the algorithm for detemining feature collision that was used.
Section 3.4 will discuss briefly the data sources used in this project, why they were chosen and specific issues with the data collection process.

@@

Programmatic collection of statistics has been neccessary for both the primary and secondary research questions, and this section presents a broad overview of the technical implementation behind the map analysis and data collection processes for both the geodata precision study and the overlap study.

@@

\subsection{Overlap study}

Once the positional error estimate was obtained from the precision study, it was used to assess the integrity of the OSM Stockholm road network.
This was done by firstly locating critical points where assigning standard widths to the roads would cause collision with other features, and then evaluating how many of those points could be eliminated by making ajustments within the found margin of error.
For the purpose of the overlap study, each road has been assigned a standard width that is in line with Swedish regulations, since in the OSM dataset, roads are represented as simple polylines, lacking width.
After the width assignment, a program iterated over all features and calculated how many of them that show overlap with any other feature, as well as the total length of the road polylines that is intersecting some other feature.
Finally, the program calculated how many of these colliding features that can be corrected by making adjustments that are within the positional accuracy of the OSM data, as obtained by the geodata precision study.

@@

Finally, when two building footprints and their vertices have been mapped, the average offset of matching vertices will be used as a measure of positional error.
The average and variance in vertex spread will be calculated and used to make an assessment of the geometric precision of the OSM dataset, and this precision will further be used to form an upper boundary on geometrical corrections that can be made in the overlap study.

\section{Overlap study}

@@

The standard road widths were obtained from a publication by the Swedish Transport Administration
\footnote{Vägars och gators utformning - Trafikverket. Publikation 2020:029. https://trafikverket.ineko.se/Files/sv-SE/71830/\\Ineko.Product.RelatedFiles/2020\_029\_vagar\_och\_gators\_utformning\_krav.pdf}
, and were presented as requirements on the minimum lane width for certain types of roads and applications.
These regulated minimum widths were used as a lower bound road with when examining the OSM dataset for potential overlap.
The upper bound was found by visual measurements of a few key locations in the Stockholm area using the Bing maps service.
Table \ref{table:road-widths} shows the minimum and maximum road widths that will be used in the overlap study.

\begin{table}[H]
    \begin{tabular}{lllll}
                      & Footpath & Residential & Secondary & Primary \\
        Minimum width & 2.0 m    & 6.0 m       & 6.5 m     & 8.0 m   \\
        Maximum width & 5.0 m    & 9.0 m       & 10.0 m    & 16.0 m
    \end{tabular}
    \captionof{table}{Standard road widths used for the overlap study}
    \label{table:road-widths}
\end{table}

@@

The algorithm begins by iterating over all road features.
Let $A$ be the current road feature.
Since this study only looks at the overlap of road features, all roads were added to a queue and gone over one by one.
The polyline of the current road feature is extracted.
Then, all other candidate features for overlap are iterated over, and their boundary polygons (for buildings) or polylines (for roads) are extracted.
Let $B$ be the current feature, road or building, that is being compared to $A$.
The algorithm then iterates over each edge $e_a$ in the polyline of $A$, and each edge $e_b$ in the polyline or polygon of $B$.
The Projected Line Segment Distance algorithm (See section 4.2.3) is used to determine the shortest distance between $e_a$ and $e_b$, and the smallest distance between any combination of edges $e_a$ and $e_b$ is taken to be the shortest distance between features $A$ and $B$.
If this distance is smaller than the minimum width of $A$, then $A$ and $B$ are colliding.

To speed up the overlap checking process, a geometric hashing solution was developed.
In this way, feature lookup could be made without comparing a feature to all other available features, but instead from a selection of the closest candidates.
The hash bucket size was 230 by 230 meters and a feature could be in multiple hash buckets.

@@

Here the results of the statistics collections will be presented.
First the results of the overlap study is presented as how many of the road features of each type that overlap with some other feature.
Then the results of the geodata precision is presented.
The final estimate of the positional accuracy of the OSM dataset will be presented, as well as the dataset completeness and statistics about building correspondence and shape accuracy.

@@

For the primary overlap study.
Table \ref{table:road-collision} shows the statistics of the overlap study, and the number of roads, number of edges and total edge length that showed intersections with other features, by road category.
For footpaths in particular, it was shown that 12.05\% of all roads collide with some other feature after extrusion to the minimum width.
The same was shown to be true for 0.11\% of the road polyline edges and 0.09\% of the total road length.
Table \ref{table:road-collision-corrected} shows the remaining roads, edges and total edges length whose intersections with other features could not be resolved by geometrical translation of a distance smaller than the positional accuracy.
For footpaths this was shown to hold for 0.51\% of all roads, 
The number of road edges and percentage of the total road length that could not be corrected by geometric translation was significantly small.

@@

    \end{tabular}
    \captionof{table}{The first result table of the overlap study. This table shows the amount of features and edges, as well as the cumulative edge length, in total and which are intersecting with any other feature.}
    \label{table:road-collision}
\end{table}

@@

    \end{tabular}
    \captionof{table}{The second result table of the overlap study. This table shows the amount of features and edges, as well as the cumulative edge length, which can be corrected by simple geometric translation.}
    \label{table:road-collision-corrected}
\end{table}

@@

Now that the colliison metrics have been obtained, conclusions may be drawn about the technical feasibility of 3D city reproduction from OSM data.
First the geodata precision study will be discussed.
The outcome of that study will be used to assess the quality of the positional accuracy estimate by looking at the dataset completeness, the low semantic accuracy and the unique characteristics of the Stockholm area.
Secondly the road overlap study will be discussed.
The main outcome is that footpaths will not be more problematic than other road types in city reproduction from OSM data.
This claim will be motivated with the data obtained in section 4.1.
Finally, the character of the feature collision cases will be discussed and the most common cases will be identified.
The chapter will conclude with a presentation of geometrical algorithms for how to correct each case automatically.
For conclusions about the usage of OSM city reproduction in engineering practise and remaining challenges, see section 5.

@@

The immediate conclusion of this study is that, in terms of feature overlap, footpaths were not significantly worse than other road categories.
As a result, it can promptly be assumed that city modelling services which already utilize OSM car road data could also use OSM footpath data without a drastic increase in feature overlap.
The biggest reason for this is thought to be that footpaths in their nature are narrow, and so narrow in fact that they can be displaced by more than half of their full width (minimum 1 meter) before the limit of half the positional accuracy (1.025 meters) is reached.
